---
title: "Writing unit tests with Hamcrest"
author: "BeDataDriven B.V."
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing unit tests with Hamcrest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

This vignette explains how to write unit tests with *hamcrest*[^1] package.

The structure of a test consists of two parts:

+ Assertion

+ Matcher

For more information about how Renjin uses *hamcrest* to write unit tests, please see the [**Using the hamcrest package to write unit tests**](http://docs.renjin.org/en/latest/writing-renjin-extensions.html#using-the-hamcrest-package-to-write-unit-tests) documentation post[^2].

## Assertions

#### assertThat

The common structure of assertThat as follows:
```{r}
assertThat({{actual}}, `matcherFun`({{expected}}))
```

where,

+ `actual`: actual object to be matched
+ `expected`: expected object to match.
+ `matcherFun`: A matcher function to check the result against the actual object.

See `Matchers` section below for available matcher calls.

There are also less central assertations such as `assertTrue` and `assertFalse` exists, and they can be used as a shortcut for `assertThat(expected, isTrue())` and `assertThat(expected, isFalse())` respectively.

## Matchers

hamcrest comes with a set of useful matchers.

#### closeTo

```{r}
assertThat(-0.50557992900139, closeTo(-0.50557, delta = 1e4))
```

#### equalTo
```{r}
assertThat(qnorm(0, 0, 1, TRUE, FALSE), equalTo(-Inf))
```

#### identicalTo
```{r}
assertThat(floor(-1.5), identicalTo(-2))
```

#### isTrue
```{r}
assertThat(is.integer(1L), isTrue())
```

#### isFalse
```{r}
assertThat(is.character(seq(10)), isFalse())
```

#### instanceOf
```{r}
res <- lm(Petal.Length ~ Petal.Width, iris)
assertThat(res, instanceOf("lm"))
```

#### deparsesTo
```{r}
assertThat(unlist(quote(sin(3.14)), recursive = FALSE), deparsesTo("sin(3.14)"))
```

#### emitsWarning
```{r}
assertThat(any(range(2.0,3.0)), emitsWarning())
```

#### throwsError
```{r}
assertThat(log("a"), throwsError())
```

## Other methods

#### not

`not` is a logical method negates the result of matcher function. It can only be used with the matcher function.

```{r}
assertThat(1, not(identicalTo(2)))
```

[^1]: *hamcrest* is a unit test framework originally developed for the Java programming language <http://hamcrest.org/>.
[^2]: Access to Renjin documentation via <http://docs.renjin.org/>.
