---
title: "Writing unit tests with Hamcrest"
author: "BeDataDriven B.V."
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Writing unit tests with Hamcrest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE, error = TRUE)
```

## Introduction

*hamcrest* is a unit test framework originally developed for Java and already
implemented in more than 30 programming languages[^1].

Although it seems that a "yet another test runner" might be too much for **R**;
however, this package is clearly intended for the developers who are familiar
with the *hamcrest* syntax and do not prefer to bother learning a new unit
testing framework such as *testthat* or *RUnit*.

Historically, *hamcrest* is extensively used in **Renjin**, a JVM-based
interpreter for **R**. Take a look at <http://www.renjin.org> for more details.

The **R** version of the *hamcrest* package provides a
*minimal* implementation from the original. 

This vignette briefly explains how to write unit tests
with *hamcrest* package for your package or project.

## Testing basics

The structure of a test consists of two parts:

### 1. Assertion

The common structure for the `assertThat()` is as follows:
```{r}
assertThat({{actual}}, `matcherFun`({{expected}}))
```

where

+ `actual`: the actual object to be matched

+ `expected`: the expected object to match

+ `matcherFun`: A matcher function to check the result against the actual object

See `Matchers` section below for the available matcher calls.

There are also secondary assertions such as `assertTrue()` and `assertFalse()`.
They can be used as a shortcut for `assertThat(expected, isTrue())` and
`assertThat(expected, isFalse())` respectively.

### 2. Matchers

hamcrest comes with a set of useful matcher functions.

| Matcher function | Returns a function that checks whether ... | Example |
|:---|:---|:---|
| `closeTo()` | its argument *is close to* an expected result within a given tolerance | `assertThat(-0.50557992900139, closeTo(-0.50557, delta = 1e4))` |
| `equalTo()` | its argument *is equal to* an expected result | `assertThat(qnorm(0, 0, 1, TRUE, FALSE), equalTo(-Inf))` |
| `identicalTo()` | its argument *is identical to* an expected result | `assertThat(floor(-1.5), identicalTo(-2))` |
| `isTrue()` | its argument is '*true*' | `assertThat(is.integer(1L), isTrue())` |
| `isFalse()` | its argument is '*false*' | `assertThat(is.character(seq(10)), isFalse())` |
| `instanceOf()` | its argument *is instance of* an expected result | `assertThat(lm(mpg ~ cyl, mtcars), instanceOf("lm"))` |
| `deparsesTo()` | its argument *deparses to* an expected result | `assertThat(unlist(quote(sin(3.14)), recursive = FALSE), deparsesTo("sin(3.14)"))`
| `emitsWarning()` | its argument *emits warning(s)* | `assertThat(any(range(2.0,3.0)), emitsWarning())`
| `throwsError()` | its argument *throws (any) error* | `assertThat(log("a"), throwsError())`
| `not()` | It is a logical method *negates* the result of matcher function. It can only be used together with a matcher function. | `assertThat(1, not(identicalTo(2)))` |

## Using hamcrest in packages

The easiest way to use `hamcrest` in your package is to create test files in the
`tests` folder where you keep all your *hamcrest* tests. For instance;

- Create a file e.g. `validation-test.R`

- Put the assertions inside the `test_hamcrest` call. This allows you to
easily distinguish the assertions, and keep those which are related to each
other together.

- Run the tests by sourcing the file such as `source("tests/validation-test.R")`.

- You can also run the individual tests inside the `test_hamcrest` block.

```{r, eval=TRUE}
library(hamcrest)
test_hamcrest("random numbers", {
  set.seed(2019)
  num <- rnorm(5)
  assertThat(num, closeTo(c(0.738522661252181,
                            -0.514760490807001,
                            -1.64018134225737,
                            0.916036784965901,
                            -1.26748197407977),
                          delta = 1e4))
})
```

When you call `R CMD check .` in the package directory, all tests will be
automatically run along with the other checks.

## Writing custom matchers

It is very possible to extend hamcrest by adding new matcher generators.
For instance;

You can write a custom matcher like this:

```{r, eval=TRUE}
isDate <- function() {
  function(actual) {
        inherits(actual, "Date")
    }
}
assertThat(Sys.Date(), isDate())
```

Or like this:

```{r, eval=TRUE}
isEvenInteger <- function() {
  function(actual) {
    actual %% 2L == 0L
  }
}
assertThat(round(pi), isEvenInteger())
```

## Hamcrest in Renjin

For more information about how **Renjin** uses *hamcrest* to write unit tests,
please see the
[**Using the hamcrest package to write unit tests**](http://docs.renjin.org/en/latest/writing-renjin-extensions.html#using-the-hamcrest-package-to-write-unit-tests)
documentation post[^2].

[^1]: Check the original hamcrest website: <http://hamcrest.org/>.
[^2]: Access to **Renjin** documentation via <http://docs.renjin.org/>.
